### 2022-06-18 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_metaweb_perspectives@8490e4857ccfb0f5bc050bd7c16f5b52dbe68c83 🚀

Treating interactions probabilistic (as opposed to binary) is a more
nuanced and realistic way to represent interactions. Dallas et al.
(2017) suggested that most links in ecological networks are cryptic,
\emph{i.e.} uncommon or hard to observe. This argument echoes Jordano
(2016): sampling ecological interactions is difficult because it
requires first the joint observation of two species, and then the
observation of their interaction. In addition, it is generally expected
that weak or rare links to be more prevalent in networks than common or
rare links (Csermely, 2004), compared to strong, persistent links; this
is notably the case in food chains, wherein many weaker links are key to
the stability of a system (Neutel et al., 2002). In the light of these
observations, we expect to see an over-representation of low-probability
interactions under a model that accurately predicts interaction
probabilities. Yet the original metaweb definition, and indeed most past
uses of metawebs, was based on the presence/absence of interactions.
Moving towards \emph{probabilistic} metawebs, by representing
interactions as Bernoulli events (see \emph{e.g.} Poisot et al., 2016),
offers the opportunity to weigh these rare interactions appropriately.
The inherent plasticity of interactions is important to capture: there
have been documented instances of food webs undergoing rapid
collapse/recovery cycles over short periods of time (\emph{e.g.}
Pedersen et al., 2017). These considerations emphasize why metaweb
predictions should focus on quantitative (preferentially probabilistic)
predictions; this should constrain the suite of appropriate models.

@@

Graph embedding (fig.~\ref{fig:embedding}) is a varied family of machine
learning techniques aiming to transform nodes and edges into a vector
space (Arsov \& Mirceva, 2019), usually of a lower dimension, whilst
maximally retaining key properties of the graph (Yan et al., 2005).
Ecological networks are an interesting candidate for the widespread
application of embeddings, as they tend to possess a shared structural
backbone (Bramon Mora et al., 2018), which hints at structural
invariants that can be revealed at lower dimensions. Indeed, food webs
are inherently low-dimensional objects, and can be adequately
represented with less than ten dimensions (Braga et al., 2019; Eklöf et
al., 2013). Simulation results by Botella et al. (2022) suggest that
there is no best method to identify architectural similarities between
networks, and that multiple approaches need to be tested and compared to
the network descriptor of interest. This matches previous, more general
results on graph embedding, which suggest that the choice of embedding
algorithm matters for the results (Goyal \& Ferrara, 2018). In
tbl.~\ref{tbl:methods}, we present a selection of common graph embedding
methods, alongside examples of their use to predict species
interactions.

@@

The popularity of graph embedding techniques in machine learning is more
than the search for structural invariants: graphs are discrete objects,
and machine learning techniques tend to handle continuous data better.
Bringing a sparse graph into a continuous, dense vector space (Xu, 2020)
opens up a broader variety of predictive algorithms, notably of the sort
that are able to predict events as probabilities (Murphy, 2022).
Furthermore, the projection of the graph itself is a representation that
can be learned; Runghen et al. (2021), for example, used a neural
network to learn the embedding of a network in which not all
interactions were known, based on nodes metadata. This example has many
parallels in ecology (see fig.~\ref{fig:prediction}), in which node
metadata can be given by phylogeny or functional traits. Rather than
directly predicting biological rules (see \emph{e.g.} Pichler et al.,
2020 for an overview), which may be confounded by the sparse nature of
graph data, learning embeddings works in the low-dimensional space that
maximizes information about the network structure. This approach is
further justified by the observation, for example, that the
macro-evolutionary history of a network is adequately represented by
some graph embeddings (RDPG; see Dalla Riva \& Stouffer, 2016).

@@

The goal of metaweb inference is to provide information about the
interactions between species at a large spatial scale. But as Herbert
(1965) rightfully pointed out, ``{[}y{]}ou can't draw neat lines around
planet-wide problems''; any inference of a metaweb at large scales must
contend with several novel, and interwoven, families of problems.

The first is the taxonomic and spatial limit of the metaweb to embed and
transfer. If the initial metaweb is too narrow in scope, notably from a
taxonomic point of view, the chances of finding another area with enough
related species (through phylogenetic relatedness or similarity of
functional traits) to make a reliable inference decreases; this would
likely be indicated by large confidence intervals during estimation of
the values in the low-rank space, or by non-overlapping trait
distributions in the known and unknown species. Alternatively a metaweb
is too large (taxonomically), then the resulting embeddings would have
interactions relative to taxonomic groups that not present in the new
location, resulting in the potential under or over estimation of the
strength of new predicted interactions. The lack of well documented
metawebs is currently preventing the development of more concrete
guidelines. The question of phylogenetic relatedness and dispersal is
notably true if the metaweb is assembled in an area with mostly endemic
species (\emph{e.g.} a system that has undergone recent radiation and
might not have an analogous system with which to draw knowledge from),
and as with every predictive algorithm, there is room for the
application of our best ecological judgement.

The second series of problems relate to determining which area should be
used to infer the new metaweb in, as this determines the species pool
that must be used. Metawebs can be constructed by assigning interactions
in a list of species within geographic boundaries. The upside of this
approach is that information at the country level is likely to be
required for biodiversity assessments, as countries set goals at the
national level (Buxton et al., 2021), and as quantitative instruments
are designed to work at these scales (Turak et al., 2017); specific
strategies are often enacted at smaller scales, nested within a specific
country (Ray et al., 2021). But there is no guarantee that these
boundaries are meaningful. In fact, we do not have a satisfying answer
to the question of ``where does a food web stop?''; the most promising
solutions involve examining the spatial consistency of network area
relationships (Fortin et al., 2021; see \emph{e.g.} Galiana et al.,
2018, 2019, 2021), which is impossible in the absence of enough
information about the network itself. This suggests that inferred
metawebs should be further downscaled to allow for \emph{a posteriori}
analyses.

The final family of problems relates less to ecological concepts and
more to the praxis of ecological research. Operating under the context
of national divisions, in large parts of the world, reflects nothing
more than the legacy of settler colonialism, which not only drive a
disparity in available ecological data, but can entrench said biases
with the machine learning models that make predictions with them.
Indeed, the use of ecological data is not an apolitical act (Nost \&
Goldstein, 2021), as data infrastructures tend to be designed to answer
questions within national boundaries, and their use often draws upon and
reinforces territorial statecraft. As per Machen \& Nost (2021), this is
particularly true when the output of ``algorithmic thinking''
(\emph{e.g.} relying on machine learning to generate knowledge) can be
re-used for governance (\emph{e.g.} enacting conservation decisions at
the national scale). We therefore recognize that predictive approaches
deployed at the continental scale, no matter their intent, originate in
the framework that contributed to the ongoing biodiversity crisis (Adam,
2014), reinforced environmental injustice (Choudry, 2013; Domínguez \&
Luoma, 2020), and \emph{e.g.} as on Turtle Island, should be replaced by
Indigenous principles of land management (Eichhorn et al., 2019; No'kmaq
et al., 2021). As we see artificial intelligence/machine learning being
increasingly mobilized to generate knowledge that is lacking for
conservation decisions (\emph{e.g.} Lamba et al., 2019; Mosebo Fernandes
et al., 2020) and drive policy decisions (Weiskopf et al., 2022), our
discussion of these tools need to go beyond the technical, and into the
governance consequences they can have.

@@

>>>>

Treating interactions as probabilistic (as opposed to binary) events is
a more nuanced and realistic way to represent them. Dallas et al. (2017)
suggested that most links in ecological networks are cryptic,
\emph{i.e.} uncommon or hard to observe. This argument echoes Jordano
(2016): sampling ecological interactions is difficult because it
requires first the joint observation of two species, and then the
observation of their interaction. In addition, it is generally expected
that weak or rare links to be more prevalent in networks than common or
rare links (Csermely, 2004), compared to strong, persistent links; this
is notably the case in food chains, wherein many weaker links are key to
the stability of a system (Neutel et al., 2002). In the light of these
observations, we expect to see an over-representation of low-probability
interactions under a model that accurately predicts interaction
probabilities. Yet the original metaweb definition, and indeed most past
uses of metawebs, was based on the presence/absence of interactions.
Moving towards \emph{probabilistic} metawebs, by representing
interactions as Bernoulli events (see \emph{e.g.} Poisot et al., 2016),
offers the opportunity to weigh these rare interactions appropriately.
The inherent plasticity of interactions is important to capture: there
have been documented instances of food webs undergoing rapid
collapse/recovery cycles over short periods of time (\emph{e.g.}
Pedersen et al., 2017). These considerations emphasize why metaweb
predictions should focus on quantitative (preferentially probabilistic)
predictions; this should constrain the suite of appropriate models.

@@

Graph embedding (fig.~\ref{fig:embedding}) is a varied family of machine
learning techniques aiming to transform nodes and edges into a vector
space (Arsov \& Mirceva, 2019), usually of a lower dimension, whilst
maximally retaining key properties of the graph (Yan et al., 2005).
Ecological networks are an interesting candidate for the widespread
application of embeddings, as they tend to possess a shared structural
backbone (Bramon Mora et al., 2018), which hints at structural
invariants that can be revealed at lower dimensions. Indeed, food webs
are inherently low-dimensional objects, and can be adequately
represented with less than ten dimensions (Braga et al., 2019; Eklöf et
al., 2013). Simulation results by Botella et al. (2022) suggest that
there is no best method to identify architectural similarities between
networks, and that multiple approaches need to be tested and compared to
the network descriptor of interest. This matches previous, more general
results on graph embedding, which suggest that the choice of embedding
algorithm matters for the results (Goyal \& Ferrara, 2018). In
tbl.~\ref{tbl:methods}, we present a selection of common graph embedding
methods, alongside examples of their use to predict species
interactions; most of these methods rely either on linear algebra, or on
pseudo-random walks on graphs.

One prominent family of approaches we do not discuss in the present
manuscript is Graph Neural Networks (GNN; \textbf{Zhou2020GraNeu?}); GNN
are, in a sense, a method to embed a graph into a dense subspace, but
belong to the family of deep learning methods, which has its own set of
practices (see \emph{e.g.} \textbf{Goodfellow2016DeeLea?}). An important
issue with methods based on deep learning is that because their
parameter space is immense, the sample size of the data fed into them
must be similarly large (typically thousands of instances). This is a
requirement for the model to converge correctly during training, but
this assumption is unlikely to be met given the size of datasets about
currently available metawebs (or single time/location species
interaction networks). This data volume requirement is mostly absent
from the techniques we list below. Furthermore, GNN still have some
challenges related to their shallow structure, and concerns related to
scalability (see \textbf{Gupta2021GraNeu?} for a review), which are
mostly absent from the methods listed in tbl.~\ref{tbl:methods}.
Assuming that the uptake of next-generation biomonitoring techniques
does indeed deliver larger datasets on species interactions
(\textbf{Bohan2017NexGlo?}), there is a potential for GNN to become an
applicable embedding/predictive technique in the coming years.

@@

The popularity of graph embedding techniques in machine learning is more
than the search for structural invariants: graphs are discrete objects,
and machine learning techniques tend to handle continuous data better.
Bringing a sparse graph into a continuous, dense vector space (Xu, 2020)
opens up a broader variety of predictive algorithms, notably of the sort
that are able to predict events as probabilities (Murphy, 2022).
Furthermore, the projection of the graph itself is a representation that
can be learned; Runghen et al. (2021), for example, used a neural
network to learn the embedding of a network in which not all
interactions were known, based on nodes metadata. This example has many
parallels in ecology (see fig.~\ref{fig:prediction}), in which node
metadata can be given by phylogeny or functional traits. Rather than
directly predicting biological rules (see \emph{e.g.} Pichler et al.,
2020 for an overview), which may be confounded by the sparse nature of
graph data, learning embeddings works in the low-dimensional space that
maximizes information about the network structure. This approach is
further justified by the observation, for example, that the
macro-evolutionary history of a network is adequately represented by
some graph embeddings (RDPG; see Dalla Riva \& Stouffer, 2016). In a
recent publication, Strydom, Bouskila, et al. (2021) have used an
embedding (based on RDPG) to project a metaweb of trophic interactions
between European mammals, and transfered this information to mammals of
Canada by using the phylogenetic distance between related clades to
infer the values in the latent sub-space into which the metaweb was
projected. By performing the RDPG step on re-constructed value, this
approach yields a probabilistic trophic metaweb for mammals of Canada
based on knowledge of European species, despite a limited (\(\approx\)
5\%) taxonomic overlap.

Graph embeddings \emph{can} serve as a dimensionality reduction method.
For example, RDPG (Strydom, Bouskila, et al., 2021) and t-SVD (Poisot et
al., 2021) typically embed networks using fewer dimensions than the
original network (the original network has as many dimensions as
species, and as many informative dimensions as trophically unique
species; Strydom, Dalla Riva, et al., 2021). But this is not
necessarilly the case -- indeed, one may perform a PCA (a special case
of SVD) to project the raw data into a subspace that improves the
efficacy of t-SNE (\textbf{Maaten2009LeaPar?}). There are many
dimensionality reductions (\textbf{Anowar2021ConEmp?}) that can be
applied to an embedded network should the need for dimensionality
reduction (for example for data visualisation) arise. In brief, many
graph embeddings \emph{can} serve as dimensionality reduction steps, but
not all do, neither do all dimensionality reduction methods provide
adequate graph embedding capacities.

@@

The goal of metaweb inference is to provide information about the
interactions between species at a large spatial scale. But as Herbert
(1965) rightfully pointed out, ``{[}y{]}ou can't draw neat lines around
planet-wide problems''; any inference of a metaweb at large scales must
contend with several novel, and interwoven, families of problems. In
this section, we list some of the most pressing research priorities
(\emph{i.e.} problems that can be adressed with subsequent data analysis
or simulations), as well as issues related to the application of these
methods at the science-policy interface.

The first open research problem is the taxonomic and spatial limit of
the metaweb to embed and transfer. If the initial metaweb is too narrow
in scope, notably from a taxonomic point of view, the chances of finding
another area with enough related species (through phylogenetic
relatedness or similarity of functional traits) to make a reliable
inference decreases; this would likely be indicated by large confidence
intervals during estimation of the values in the low-rank space, meaning
that the representation of the original graph is difficult to transfer
to the new problem. In addition, other problems can arise due to
non-overlapping trait distributions in the known and unknown species.
Alternatively a metaweb is too large (taxonomically), then the resulting
embeddings would need to represent interactions between taxonomic groups
that are not present in the new location. This would lead to a much
higher variance in the starting dataset, and to under-dispersal in the
target dataset, resulting in the potential under or over estimation of
the strength of new predicted interactions. The lack of well documented
metawebs is currently preventing the development of more concrete
guidelines. The question of phylogenetic relatedness and dispersal is
notably true if the metaweb is assembled in an area with mostly endemic
species (\emph{e.g.} a system that has undergone recent radiation and
might not have an analogous system with which to draw knowledge from),
and as with every predictive algorithm, there is room for the
application of our best ecological judgement. Because this problem
relates to dispersal of species in the geographic or phylogenetic space,
it can certainly be approached through assessing the performance of
embedding transfer in simulated starting/target species pools.

The second series of problems relate to determining which area should be
used to infer the new metaweb in, as this determines the species pool
that must be used. Metawebs can be constructed by assigning interactions
in a list of species within geographic boundaries. The upside of this
approach is that information at the country level is likely to be
required for biodiversity assessments, as countries set goals at the
national level (Buxton et al., 2021), and as quantitative instruments
are designed to work at these scales (Turak et al., 2017); specific
strategies are often enacted at smaller scales, nested within a specific
country (Ray et al., 2021). But there is no guarantee that these
boundaries are meaningful. In fact, we do not have a satisfying answer
to the question of ``where does a food web stop?''; the most promising
solutions involve examining the spatial consistency of network area
relationships (Fortin et al., 2021; see \emph{e.g.} Galiana et al.,
2018, 2019, 2021), which is impossible in the absence of enough
information about the network itself. This suggests that inferred
metawebs should be further downscaled to allow for \emph{a posteriori}
analyses. The methodology for metaweb downscaling is currently limited,
and it is likely that the sustained effort to characterize the spatial
dependency of food web structure will lead to more prescriptive
guidelines about the need for prediction downscaling.

The final family of problems relates less to ecological methods than to
the praxis of ecological research. Operating under the context of
national divisions, in large parts of the world, reflects nothing more
than the legacy of settler colonialism, which drives a disparity in
available ecological data. Applying any embedding to biased data does
not debias them, but instead embeds these very same biases, propagating
them to the machine learning models using embeddings tomake predictions.
Indeed, the use of ecological data is not an apolitical act (Nost \&
Goldstein, 2021), as data infrastructures tend to be designed to answer
questions within national boundaries (therefore placing contingencies on
what is available to be embedded), and their use often draws upon and
reinforces territorial statecraft. As per Machen \& Nost (2021), this is
particularly true when the output of ``algorithmic thinking''
(\emph{e.g.} relying on machine learning to generate knowledge) can be
re-used for governance (\emph{e.g.} enacting conservation decisions at
the national scale). As information on species interaction networks
structure is increasingly leveraged as a tool to guide conservation
actions (see \emph{e.g.} recent discussions for food-web based
conservation; \textbf{Eero2021UseFoo?}; \textbf{NamanFooWeb?};
\textbf{Stier2017IntExp?}), the need to appraise and correct biases that
are unwittingly propagated to algorithms when embedded from the original
data is paramount. Predictive approaches deployed at the continental
scale, no matter their intent, originate in the framework that
contributed to the ongoing biodiversity crisis (Adam, 2014) and
reinforced environmental injustice (Choudry, 2013; Domínguez \& Luoma,
2020). Particularly on Turtle Island and other territories that were
traditionally stewarded by Indigenous people, these approaches should be
replaced (or at least guided and framed) by Indigenous principles of
land management (Eichhorn et al., 2019; No'kmaq et al., 2021), as part
of an ``algorithm-in-the-loop'' approach. Human-algorithm interactions
are notoriously difficult and can yield adverse effect
(\textbf{Green2019DisInt?}; \textbf{Stevenson2021AlgRis?}), suggesting
the need to systematically study them for the specific purpose of
biodiversity governance, as well as to improve the algorithmic literacy
of decision makers. As we see artificial intelligence/machine learning
being increasingly mobilized to generate knowledge that is lacking for
conservation decisions (\emph{e.g.} Lamba et al., 2019; Mosebo Fernandes
et al., 2020) and drive policy decisions (Weiskopf et al., 2022), our
discussion of these tools need to go beyond the technical and
statistical, and into the governance consequences they can have.

@@

\leavevmode\hypertarget{ref-Strydom2021SvdEnt}{}%
Strydom, T., Dalla Riva, G. V., \& Poisot, T. (2021). SVD Entropy
Reveals the High Complexity of Ecological Networks. \emph{Frontiers in
Ecology and Evolution}, \emph{9}.
\url{https://doi.org/10.3389/fevo.2021.623141}
